Measuring the double-drift illusion and its resets with hand trajectories

### 2022-01-21 Manuscript.Rmd
final round of manuscript edits before 3rd submission

We compare the illusory direction of the heading against the predictions of a simple vector combination model of perceived direction from Tse and Hsieh [-@Tse2006], Cavanagh and Tse [-@Cavanagh2019] and Heller et al. [-@Heller2021]. A more sophisticated model allows predicting perceived speed as well [@Johnston2013], but this model predicts perceived direction and speed for a mixture of first order motions, and it is unclear if this works for a combination of first and second order motion (what we call internal and external speed here). In the vector average model we use here, the perceived direction

To determine if hand tracking produces an illusion similar to that seen in other studies, we compare our results to those reported by Gurnsey and Biard [-@Gurnsey2012], Cavanagh and Tse [-@Cavanagh2019] and Heller et al. [-@Heller2021] who all used similar displays. We also compare the illusory direction of the heading against the predictions of a simple vector combination model [@Tse2006; @Cavanagh2019; @Heller2021]. There are alternative models that could be considered, one from Kwon et al. [-@Kwon2015] and another from Johnston and Scarfe [-@Johnston2013]. Kwon et al.’s [-@Kwon2015] Kalman filter tracking model assumes a saturation of the position shifts by 200 ms, something we do not observe here. Johnson and Scarfe’s [-@Johnston2013] model used a harmonic average to combine vectors, a process that gives more weight to vectors with lower magnitudes. This model was used to estimate the perceived direction of an array of stationary Gabors with various internal motions but is less suited to the double-drift stimulus that has only two vectors, one of which is a second-order motion (the motion of the Gabor itself). In the simple vector combination model we use here, the perceived direction is a vector combination of the external ($V_e$) and internal ($V_i$) motions weighted by a constant $K$. That is, the deviation from the physical direction is given by:

@@

The X coordinates of the detected reset points give the deviation from the vertical path and are used as the spatial offset of the reset. We also need the time of the reset. The gabor moves at a uniform speed along the Y axis so distance along Y denotes the elapsed time for the physical path. However, the illusory path of the gabor does not have to stay in alignment with its physical location on the Y axis. If it did, its speed and path length would have to increase for larger illusion angles: for example, 41% faster and farther for a 45° illusion, 73% faster and farther for a 60° illusion. Observers have not reported these increases in speed or path length compared to the control with no internal motion. Moreover, in a recent unpublished study we directly measured the perceived path length of the illusion. The path length was always fixed proportion of the physical path (around 70%) for all illusion strengths, Gabor speeds and physical path lengths.  This means that the perceived speed must be a fixed proportion of the physical speed as well and therefore, in the tracings, it is the distance along the traced path (the radius in our diagrams) that tells us the time since the trial began — not the distance along the physical path (the Y axis). For completeness, we also analyze the distributions along the Y axis.

@@

While the marginal distribution plot in Fig 5A qualitatively appear an equally good fit to the data, the better fit for the distribution on the time of resets could be explained be a better generalization throughout the data set. We test this by dividing the data into segments, each with an equal number of resets, based on variables orthogonal to the fitted variable (Y-coordinates for the spatial offset of resets and angle for reset time). The median for each segment of data is shown in Fig 5B,C. We fit a gamma distribution to each segment, then use the data points outside the segment to get a log likelihood as a measure of generalization and sum them (spatial: -3087.9, temporal: -1806.3). We use this to calculate AICs and a relative likelihood (spatial: 6191.7, temporal: 3628.6, _p_<.001) which shows that a gamma distribution fit to reset times in one segment of the data generalizes better to the rest of the data, as compared to a gamma distribution fit to the spatial offset of resets.

@@

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in their experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe [-@Nakayama2020] claim that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. However, the distribution of resets throughout time are largely in line with a mechanism based on random distractions of attention, which may involve eye-movement planning and execution. The rate of resets that we found here, ~0.75 / _s_, is too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements. On the other hand, gamma distributions [e.g. @McGill1965; @Wolfe2002] do model a random occurrence of events (rate parameter) where only some fraction of these trigger an action (shape parameter). In other words, the good fit of a gamma distribution may reveal some properties of the process underlying resets of the double-drift illusion. While the process giving rise to resets is likely memoryless, a process giving rise to a gamma distribution (with shape > 1) is not, it does capture two ideas: 1) resets are triggered by random events (such as distractions of attention or microsaccades), and 2) only some of these actually lead to a reset. An alternative to the gamma distribution, the ex-Gaussian [e.g. @Matzke2009; @Guy2020], could be a more realistic alternative but does not fit the data as well. However, the process that actually leads to resets remains to be seen in future work.

>>>>

To determine if hand tracking produces an illusion similar to that seen in other studies, we first compare our results to those reported by Heller et al. [-@Heller2021] who used similar displays. We also compare the illusory direction of the heading against the predictions of a simple vector combination model [@Tse2006; @Cavanagh2019; @Heller2021]. There are alternative models that could be considered, one from Kwon et al. [-@Kwon2015] and another from Johnston and Scarfe [-@Johnston2013]. Kwon et al.’s [-@Kwon2015] Kalman filter tracking model assumes a saturation of the position shifts by 200 ms, something we do not observe here. Johnson and Scarfe’s [-@Johnston2013] model used a harmonic average to combine vectors, a process that gives more weight to vectors with lower magnitudes. This model was used to estimate the perceived direction of an array of stationary Gabors with various internal motions but is less suited to the double-drift stimulus that has only two vectors, one of which is a second-order motion (the motion of the Gabor itself). In the simple vector combination model we use here, the perceived direction is a vector combination of the external ($V_e$) and internal ($V_i$) motions weighted by a constant $K$. That is, the deviation from the physical direction is given by:

@@

The X coordinates of the detected reset points give the deviation from the vertical path and are used as the spatial offset of the reset. We also need the time of the reset. The gabor moves at a uniform speed along the Y axis so distance along Y denotes the elapsed time for the physical path. However, the illusory path of the gabor does not have to stay in alignment with its physical location on the Y axis. If it did, its speed and path length would have to increase for larger illusion angles: for example, 41% faster and farther for a 45° illusion, 73% faster and farther for a 60° illusion. Observers have not reported these increases in speed or path length compared to the control with no internal motion. Moreover, in a recent unpublished study we directly measured the perceived path length of the illusion. The path length was always fixed proportion of the physical path (around 70%) for all illusion strengths, Gabor speeds and physical path lengths. This means that the perceived speed along the illusory path must be a fixed proportion of the physical speed as well and therefore, in the tracings, it is the distance along the traced path (the radius in our diagrams) that tells us the time since the trial began — not the distance along the physical path (the Y axis).

@@

While the marginal distribution plot in Fig 5A qualitatively appear an equally good fit to the data, the fit for the distribution on the time of resets maintains a better fit throughout the data set. We demonstrate this by dividing the data into segments, each with an equal number of resets, based on variables orthogonal to the fitted variable (Y-coordinates for the spatial offset of resets and angle for reset time). The median for each segment of data is shown in Fig 5B,C. We fit a gamma distribution to each segment, then use the data points outside the segment to get a log likelihood as a measure of generalization and sum them. We use this to calculate AICs and a relative likelihood (spatial: 6191.7, temporal: 3628.6, _p_<.001) which shows that a gamma distribution fit to reset times in one segment of the data generalizes better to the rest of the data, as compared to a gamma distribution fit to the spatial offset of resets.

@@

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in their experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe [-@Nakayama2020] claim that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. However, the distribution of resets throughout time are roughly in line with a mechanism based on random distractions of attention, which may involve eye-movement planning and execution. The rate of resets that we found here, ~0.75 / _s_, is too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements. On the other hand, gamma distributions [e.g. @McGill1965; @Wolfe2002] do model a random occurrence of events (rate parameter) where only some fraction of these trigger an action (shape parameter). In other words, the good fit of a gamma distribution may reveal some properties of the process underlying resets of the double-drift illusion. While the process giving rise to resets is likely memoryless, a process giving rise to a gamma distribution (with shape > 1) is not. Nevertheless, it does capture two ideas: 1) resets are triggered by random events (such as distractions of attention or microsaccades), and 2) only some of these actually lead to a reset. An alternative to the gamma distribution, the ex-Gaussian [e.g. @Matzke2009; @Guy2020], could be a more realistic alternative but does not fit the data as well. In any case, the process that actually leads to resets remains to be seen in future work.
